---
title: "INFO510FA24-AB-3"
author: "Aditya Bandimatt"
date: "2024-11-27"
output: html_document
header-includes:  
 - \usepackage[utf8]{inputenc}
 - \usepackage{amsmath} 
 - \usepackage{unicode-math}
 - \setmathfont{XITS Math}
---
### Gibbs Sampling
#### Setup and load libraries  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
```

Generate some synthetic data from a normal distribution, which we will analyze
using our Bayesian model.

```{r}
# Set seed for reproducibility
set.seed(123)

# Generate synthetic data from N(5, 2^2)
n <- 100 # Number of observations
true_mu <- 5
true_sigma <- 2
y <- rnorm(n, mean = true_mu, sd = true_sigma)

# Plot the data
ggplot(data.frame(y), aes(x = y)) +
  geom_histogram(binwidth = 0.5, color = "black", fill = "lightblue") +
  ggtitle("Histogram of the Observed Data") +
  xlab("y") + ylab("Frequency")

```

### setting the priors
$$\mu \sim N(\mu_0 = 0, \tau^2 = 10^2)$$
$$\sigma^2 \sim IG(a = 2, b = 2)$$

```{r}
# Prior parameters
mu_0 <- 0
tau2 <- 10^2
a <- 2
b <- 2

```

## Gibbs Sampling
#Sampling mu
$$
\mu \mid \sigma^2, y \sim N\left( \frac{\tau^{-2} \mu_0 + n \bar{y}}{\tau^{-2} + n}, \frac{1}{\tau^{-2} + n} \right)
 $$
#sampling sigma^2
$$
\sigma^2 \mid \mu, y \sim IG\left(a + \frac{n}{2}, b + \frac{1}{2} \sum_{i=1}^{n} (y_i - \mu)^2 \right)
$$

```{r}
# Gibbs sampler function
gibbs_sampler <- function(y, mu_0, tau2, a, b, n_iter = 10000) {
  n <- length(y)
  y_bar <- mean(y)
  mu_samples <- numeric(n_iter)
  sigma2_samples <- numeric(n_iter)

  # Initialize parameters
  mu <- mean(y)
  sigma2 <- var(y)

  for (iter in 1:n_iter) {
    # Sample mu from its conditional posterior
    mu_n <- (mu_0 / tau2 + n * y_bar / sigma2) / (1 / tau2 + n / sigma2)
    sigma_mu_n <- sqrt(1 / (1 / tau2 + n / sigma2))
    mu <- rnorm(1, mean = mu_n, sd = sigma_mu_n)

    # Sample sigma^2 from its conditional posterior
    shape <- a + n / 2
    rate <- b + sum((y - mu)^2) / 2
    sigma2 <- 1 / rgamma(1, shape = shape, rate = rate)

    # Store samples
    mu_samples[iter] <- mu
    sigma2_samples[iter] <- sigma2
  }

  return(list(mu_samples = mu_samples, sigma2_samples = sigma2_samples))
}

```
```{r}
# Run Gibbs sampler
n_iter <- 10000
samples <- gibbs_sampler(y, mu_0, tau2, a, b, n_iter)


# we have 10,000 samples from the posterior distributions ofðœ‡and Ïƒ

#Let's visualize the posterior distributions and provide summary statistics

# Posterior samples
mu_samples <- samples$mu_samples
sigma2_samples <- samples$sigma2_samples

# Plot posterior distribution of mu
ggplot(data.frame(mu_samples), aes(x = mu_samples)) +
  geom_histogram(binwidth = 0.05, color = "black", fill = "lightblue") +
  ggtitle("Posterior Distribution of Mu") +
  xlab("Mu") + ylab("Frequency")

```
```{r}
# Plot posterior distribution of sigma^2
ggplot(data.frame(sigma2_samples), aes(x = sigma2_samples)) +
  geom_histogram(binwidth = 0.1, color = "black", fill = "lightgreen") +
  ggtitle("Posterior Distribution of Sigma^2") +
  xlab("Sigma^2") + ylab("Frequency")

```

```{r}
# Summary statistics
mu_mean <- mean(mu_samples)
mu_ci <- quantile(mu_samples, c(0.025, 0.975))
sigma2_mean <- mean(sigma2_samples)
sigma2_ci <- quantile(sigma2_samples, c(0.025, 0.975))

mu_mean
mu_ci
sigma2_mean
sigma2_ci

```

#### Explanation of Results
- The **posterior distribution of mu** (the population mean) gives us an
estimate of where the true mean lies, accounting for both the data and our
prior beliefs. The 95% credible interval is an interpretation of uncertainty,
meaning we're 95% confident that the true mean is within this range
- The **posterior distribution of sigma^2** (the population variance) also
shows uncertainty, helping us estimate the range of the variance in the
population.
